---
output: github_document
---

# Preliminary Damage Assessments for Presidential Disaster Declaration Requests

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  fig.path = "man/figures/README-",
  dpi = 300,
  eval = TRUE,
  echo = TRUE)

library(preliminarydamages)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(urbnthemes)

set_urbn_defaults(style = "print")

options(scipen = 999)
```

## Disclaimer

The structured data returned by this package are created by extracting values
from PDF reports. This process is inherently imprecise, and it is likely that
some values in the structured data are incorrect relative to the source PDFs. 
We have taken significant steps to mitigate the prevalence of such errors through
both manual spot-checking and programmatic testing, but users should take this into 
account if they choose to use these data. 

## Installation

```{r install, eval = FALSE}
## install.packages("remotes")
remotes::install_github("UrbanInstitute/preliminary-damage-assessments")
```

## Overview

When a state, territory, or tribal (STT) government believes an event exceeds its
capacity to respond, it may request a major disaster declaration from the
president. Before that request is granted or denied, FEMA and its partners
conduct a *preliminary damage assessment* (PDA) that estimates the scale of
the damage and the characteristics of the communities impacted. Information 
collected during PDAs does not determine whether a disaster declaration request
is approved--that is ultimately at the sole discretion of the president--but
it does form an important basis for whether the Federal Emergency Management Agency (FEMA) 
recommends to the president that the president declare a disaster.

STT governments are the leads on disaster declaration requests, and FEMA is the primary
agency that coordinates this process, including participating in conducting joint 
preliminary damage assessments with the STT and local governments and producing the summative
PDA reports located at https://www.fema.gov/disaster/how-declared/preliminary-damage-assessments/reports.
These reports consistently document key attributes reflecting the PDA findings, such as total
estimated Public Assistance (PA) and Individual Assistance (IA) costs, as well as the extent of
damage to housing and the characteristics of impacted populations. 

However, because these reports are text-based PDFs, and because FEMA does not publish
these data in a tabular format, it has been challenging to date to systematically 
access and use this valuable information to understand when and under what conditions 
disaster declaration requests are approved and denied. 

`get_preliminary_damage_assessments()` extracts these data from the
PDA reports FEMA publishes and then joins them to authoritative datasets from
FEMA describing disaster declaration request approvals and denials.

## Data sources

| Role | Dataset | Access |
|---|---|---|
| Damage estimates | PDA reports | `preliminarydamages::get_preliminary_damage_assessments()` |
| Approved requests | `DisasterDeclarationsSummaries` | OpenFEMA, via `rfema::open_fema()` |
| Denied requests | `DeclarationDenials` | OpenFEMA, via `rfema::open_fema()` |

The two FEMA datasets are *authoritative*: they are FEMA's own structured
records of which declaration requests were approved and denied. The PDAs, by contrast,
are text-extracted from unstructured PDFs and carry the caveats described in
`?get_preliminary_damage_assessments`. We treat observations
from the authoritative FEMA datasets as defining the universe of declarations,
and we map PDAs back to this universe so that we can better understand the 
relationships between preliminary damages and disaster declaration decisionmaking. 

## PDA Data Preparation

The package provides three functions for accessing PDA data: 

 - `scrape_pda_pdfs()` finds the URLs to every PDA report at 
  https://www.fema.gov/disaster/how-declared/preliminary-damage-assessments/reports and
  then systematically downloads these PDFs into a user-specified directory. These PDFs
  are the basis for the dataset produced by the following function. Because FEMA receives 
  and the President makes disaster request decisions on an ongoing basis, the results of this
  function quickly become stale: new PDA reports are available online but are not reflected
  in the user's specified directory. Accordingly, it is important to re-run this function 
  frequently if the most up-to-date data are important. The function does not re-download reports
  that are already saved in the user-specified directory, so after the initial run, which can be 
  slow, subsequent runs should be very fast--they only add new reports.
- `get_preliminary_damage_assessments()` walks the user's specified directory and extracts 
  structured attributes from each PDA report, compiling data from all reports in the directory 
  into a consistent dataset, and joining PDA attributes back to the authoritative FEMA-published 
  datasets of declaration request approvals and denials.
- `transform_pda_counties()` restructures the request-level data from `get_preliminary_damage_assessments()`
  into a county-request structured dataset. While requests originate from STT governments, they 
  delineate specific counties--both estimated costs per county, and the programs (PA and IA) that
  are requested to be made available for each county. 

```{r load-pdas, eval = FALSE}
## run this to update (or create for the first time) the local 
## store of all PDFS
scrape_pda_pdfs(cache_directory = file.path("data", "pdfs"))
```

```{r}
## this extracts attributes from all reports
## use_cache = TRUE will not capture newly-scraped PDFs
pdas = get_preliminary_damage_assessments(
  file_path = file.path("data", "pdas.csv"),
  directory_path = file.path("data", "pdfs"),
  use_cache = TRUE)
```

## Analysis

With our data in hand, it's trivial to summarize the universe of disaster declaration requests
over time:

```{r declarations-over-time}
percent_approved = pdas %>%
  count(fema_decision) %>%
  mutate(percent = round(prop.table(n) * 100, digits = 0)) %>%
  filter(fema_decision == "Approved") %>%
  pull(percent)

pdas %>%
  ggplot() +
  geom_bar(aes(x = fema_decision_year, fill = fema_decision)) +
  scale_x_continuous(breaks = seq(1950, 2030, by = 10)) +
  labs(
    title = str_c(glue::glue("Since the 1950s, {percent_approved}% of disaster ", 
    "declaration requests have been approved")),
    x = "",
    y = "")
```

But access to systematic information on damages and housing impacts are the novel 
contributions from the PDAs. 

One of the primary factors intended to determine whether PA is declared is whether the estimated 
per-capita damages (in dollars), per person, across the entire STT government, 
exceed an established per-capita threshold or indicator, known as the (statewide) 
Per Capita Impact (PCI) indicator. 

In 2026, this PCI is $1.94. There is also a separate PCI at the county level, which accounts 
for the fact that disaster damages could be concentrated in a subset of counties and warrant a 
declaration for one or more counties, even if total damages do not meet the statewide PCI and 
a statewide disaster declaration may not be appropriate. The countywide PCI in 2026 is $4.86.

(Because published PDAs are only available from ~2008 onward, we restrict the data 
to that window.)

```{r pa-threshold-ratio}
pdas %>%
  filter(pda_matched == 1, pda_pa_requested == 1, fema_decision_year > 2007) %>%
  ggplot(aes(x = fema_decision, y = log(pda_pa_threshold_ratio), color = fema_decision)) +
    geom_boxplot(fill = NA) +
    geom_hline(yintercept = 0) +
    annotate(
      "text", x = 1.5, y = 2.75, hjust = 1,
      label = str_wrap("Requests above zero exceed the per-capita threshold", 25)) +
    annotate(
      "curve", x = 1.4, y = 2.1, xend = 1.5, yend = 0.1, curvature = -0.35,
      arrow = arrow(length = unit(0.02, "npc"), type = "closed")) +
    labs(
      title = str_c(
        "Many denied requests exceed the per-capita damages threshold") %>% str_wrap(100),
        subtitle = str_c(
          "Ratio of Public Assistance per-capita damages relative to the statewide ",  
          "per-capita indicator, logged") %>% str_wrap(100),
      x = "",  y = "")
```

## Use of Generative AI

This package was developed in part using agentic AI, under the supervision of the author. 