---
output: github_document
---

# Preliminary Damage Assessments for Presidential Disaster Declaration Requests

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  fig.path = "man/figures/README-",
  dpi = 300,
  eval = TRUE,
  echo = TRUE)

library(preliminarydamages)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(urbnthemes)

set_urbn_defaults(style = "print")

options(scipen = 999)
```

## Data disclaimer

The structured data returned by this package are created by extracting values
from PDF reports. This process is inherently imprecise, and it is likely that
some values in the structured data are incorrect relative to the source PDFs. 
We have taken significant steps to mitigate the prevalence of such errors through
both manual spot-checking and programmatic testing, but users should take this into 
account if they choose to use these data. 

Further, this package is under active development, and will continue to be updated to
make the returned fields and specific values more comprehensive and robust. The data
include a parser version number that reflects the vintage of the codebase used to produce
the data; this can be a helpful tool if you want to pin your results to a specific version
of the code. Using a dependency manager such as `renv` can help ensure that you're able to
record and refer back to the specific GitHub commit from which you installed the package. 

## Installation

```{r install, eval = FALSE}
## install.packages("remotes")
remotes::install_github("UrbanInstitute/preliminary-damage-assessments")
```

## Overview

When a state, territory, or tribal (STT) government believes an event exceeds its
capacity to respond, it may request a major disaster declaration from the
president. Before that request is granted or denied, FEMA and its partners
conduct a *preliminary damage assessment* (PDA) that estimates the scale of
the damage and the characteristics of the communities impacted. Information 
collected during PDAs does not determine whether a disaster declaration request
is approved--that is ultimately at the sole discretion of the president--but
it does form an important basis for whether the Federal Emergency Management Agency (FEMA) 
recommends to the president that the president declare a disaster.

STT governments are the leads on disaster declaration requests, and FEMA is the primary
agency that coordinates this process, including participating in conducting joint 
preliminary damage assessments with the STT and local governments and producing the summative
PDA reports located at https://www.fema.gov/disaster/how-declared/preliminary-damage-assessments/reports.
These reports consistently document key attributes reflecting the PDA findings, such as total
estimated Public Assistance (PA) and Individual Assistance (IA) costs, as well as the extent of
damage to housing and the characteristics of impacted populations. 

However, because these reports are text-based PDFs, and because FEMA does not publish
these data in a tabular format, it has been challenging to date to systematically 
access and use this valuable information to understand when and under what conditions 
disaster declaration requests are approved and denied. 

`get_preliminary_damage_assessments()` extracts these data from the
PDA reports FEMA publishes and then joins them to authoritative datasets from
FEMA describing disaster declaration request approvals and denials.

## Data sources

| Role | Dataset | Access |
|---|---|---|
| Damage estimates | PDA reports | `preliminarydamages::get_preliminary_damage_assessments()` |
| Approved requests | `DisasterDeclarationsSummaries` | OpenFEMA, via `rfema::open_fema()` |
| Denied requests | `DeclarationDenials` | OpenFEMA, via `rfema::open_fema()` |

The two FEMA datasets are *authoritative*: they are FEMA's own structured
records of which declaration requests were approved and denied. The PDAs, by contrast,
are text-extracted from unstructured PDFs and carry the caveats described in
`?get_preliminary_damage_assessments`. We treat observations
from the authoritative FEMA datasets as defining the universe of declarations,
and we map PDAs back to this universe so that we can better understand the 
relationships between preliminary damages and disaster declaration decisionmaking. 

## Data and analysis notes

The disaster declaration process is complex and nuanced. Users of these data should familiarize themselves
with the process -- at a minimum, by reading FEMA's resources at https://www.fema.gov/disaster/how-declared --
before beginning an analysis. Some important considerations relevant to how these data are created and can be used:

 - Tribal nations that are lead applicants on a request for a declaration navigate a different
   process with different evaluation criteria. The process shares many similarities with the 
   process for a state- or territory-led request, but it may be more appropriate to analyze 
   tribe-led requests separately. https://www.urban.org/urban-wire/how-tribes-can-pursue-disaster-declarations-and-federal-recovery-funding
 - Funding programs are authorized discretely and for specific geographies. 
   A declaration could approve Public Assistance and deny a request for Individual Assistance, 
   or could authorize Public Assistance but only for a subset of the counties for which it was requested. 
   Further, some PDAs refer to other disaster declaration-related programs, including the Hazard Mitigation Grant Program,
   the Small Business Association's Disaster Loans, or particular programs within the umbrella Individual 
   Assistance program. This package only derive data for specific programs that are frequently and consistently
   referred to across PDA reports.  
 - County-level data is derived from a standard, Public Assistance-specific field in the PDA report template. 
   `transform_pda_counties` produces observations about counties for which Public Assistance was requested,
   if it was requested and if counties were itemized. Do not attempt to use county-level observations to make
   claims about other funding programs.
 - As noted above, 

## Data preparation

The package provides three core functions for accessing PDA data: 

 - `scrape_pda_pdfs()` finds the URLs to every PDA report at 
  https://www.fema.gov/disaster/how-declared/preliminary-damage-assessments/reports and
  then systematically downloads these PDFs into a user-specified directory. These PDFs
  are the basis for the dataset produced by the following function. Because FEMA receives 
  and the President makes disaster request decisions on an ongoing basis, the results of this
  function quickly become stale: new PDA reports are available online but are not reflected
  in the user's specified directory. Accordingly, it is important to re-run this function 
  frequently if the most up-to-date data are important. The function does not re-download reports
  that are already saved in the user-specified directory, so after the initial run, which can be 
  slow, subsequent runs should be very fast--they only add new reports. `check_pda_cache()` is a
  simple utility to help gauge whether your cache is approximately current with full universe of PDA
  reports on FEMA's website.
- `get_preliminary_damage_assessments()` walks the user's specified directory and extracts 
  structured attributes from each PDA report, compiling data from all reports in the directory 
  into a consistent dataset, and joining PDA attributes back to the authoritative FEMA-published 
  datasets of declaration request approvals and denials.
- `transform_pda_counties()` restructures the request-level data from `get_preliminary_damage_assessments()`
  into a county-request structured dataset. While requests originate from STT governments, they 
  delineate specific counties---both estimated costs per county, and the programs (PA and IA) that
  are requested for each county. `pda_county_indicator` and `fema_county_indicator` record which 
  source named each county. When a request has no county-level information, the county columns are `NA`. 
  For statewide requests, a record is returned for each county in the state, even when counties are not
  named individually. Every county row carries `fema_declaration_request_number`, FEMA's own
  identifier for the request, so county-level results can be grouped or joined back to their event.

```{r load-pdas, eval = FALSE}
## run this to update (or create for the first time) the local 
## store of all PDFS
scrape_pda_pdfs(cache_directory = file.path("data", "pdfs"))
```

```{r}
## this extracts attributes from all PDA reports in directory_path
## use_cache = TRUE will not capture newly-scraped PDFs
pdas = get_preliminary_damage_assessments(
  file_path = file.path("data", "pdas.csv"),
  directory_path = file.path("data", "pdfs"),
  use_cache = TRUE)
```

## Disaster-Level Analysis

With our data in hand, it's trivial to summarize the universe of disaster declaration requests
over time:

```{r declarations-over-time}
percent_approved = pdas %>%
  count(fema_decision) %>%
  mutate(percent = round(n / sum(n) * 100, digits = 0)) %>%
  filter(fema_decision == "Approved") %>%
  pull(percent)

pdas %>%
  ggplot() +
  geom_bar(aes(x = fema_decision_year, fill = fema_decision)) +
  scale_x_continuous(breaks = seq(1950, 2030, by = 10)) +
  labs(
    title = glue::glue("Since the 1950s, {percent_approved}% of disaster ", 
    "declaration requests have been approved"),
    x = "",
    y = "")
```

But access to systematic information on damages and housing impacts are the novel 
contributions from the PDAs. 

One of the primary factors intended to determine whether PA is declared is whether the estimated 
per-capita damages (in dollars), per person, across the entire STT government, 
exceed an established per-capita threshold or indicator, known as the (statewide) 
Per Capita Impact (PCI) indicator. In 2026, this PCI is $1.94. 

(Because published PDAs are only available from ~2008 onward, we restrict the data 
to that window.)

```{r pa-threshold-ratio}
pdas %>%
  filter(pda_matched == 1, pda_pa_requested == 1, fema_decision_year > 2007) %>%
  ggplot(aes(x = fema_decision, y = log(pda_pa_threshold_ratio), color = fema_decision)) +
    geom_boxplot(fill = NA) +
    geom_hline(yintercept = 0) +
    annotate(
      "text", x = 1.5, y = 2.75, hjust = 1,
      label = str_wrap("Requests above zero exceed the per-capita threshold", 25)) +
    annotate(
      "curve", x = 1.4, y = 2.1, xend = 1.5, yend = 0.1, curvature = -0.35,
      arrow = arrow(length = unit(0.02, "npc"), type = "closed")) +
    labs(
      title = str_c(
        "Many denied requests exceed the per-capita damages threshold") %>% str_wrap(100),
        subtitle = str_c(
          "Ratio of Public Assistance per-capita damages relative to the statewide ",  
          "per-capita indicator, logged") %>% str_wrap(100),
      x = "",  y = "")
```

## County-Level Analysis

There is also a separate PCI at the county level, which accounts for the fact that disaster damages 
could be concentrated in a subset of counties and warrant a declaration for one or more counties, 
even if total damages do not meet the statewide PCI and a statewide disaster declaration may not be 
appropriate. The countywide PCI in 2026 is $4.86.

How well does a PDA capture the counties that end up in a declaration? FEMA publishes
the counties it designated only for approved requests, so here we compare PDA-listed counties
against FEMA-published counties in approved requests. PDAs only systematically list the counties
assessed for Public Assistance, so we likewise limit our comparison on that axis. 

The two sources can differ for a few reasons: a declaration is often amended
after the initial PDA report is published to add counties, and a county can be assessed (and thus
included in a PDA report) without being designated to receive funding--what is reflected in the 
authoritative FEMA data.

We can see that coverage is generally fairly high, with nearly 80% correspondence between authoritative
FEMA data and PDA reports. 

```{r pda-fema-county-agreement}
county_pdas = pdas %>% transform_pda_counties()

## here we look at county-level correspondence 
county_coverage = county_pdas %>%
  filter(pda_matched, fema_decision == "Approved") %>%
  filter(
    .by = fema_declaration_request_number,
    any(pda_county_indicator, na.rm = TRUE)) %>%
  ## only places with a county code can be matched between the two sources
  filter(!is.na(pda_county_geoid)) %>%
  ## a report covering two incidents lists some counties twice
  distinct(fema_declaration_request_number, pda_county_geoid, .keep_all = TRUE) %>%
  mutate(
    fema_pa_designated = coalesce(fema_county_indicator & fema_pa_declared_county, FALSE)) %>%
  ## counties FEMA designated only for other programs are outside the comparison
  filter(pda_county_indicator | fema_pa_designated)

county_coverage  %>%
  count(pda_county_indicator, fema_pa_designated) %>%
  mutate(percent = n / sum(n))

disaster_coverage = county_coverage %>%
  filter(!(pda_county_indicator & fema_pa_designated)) %>%
  count(fema_disaster_number, sort = TRUE) %>%
  mutate(
    `# of disasters` = row_number(),
    running_sum = cumsum(n),
    running_percent = running_sum / sum(n))

disaster_coverage  %>%
  ggplot() +
    geom_line(aes(x = `# of disasters`, y = running_percent)) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
      title = "Under 5% of all disasters account for half of all county non-correspondence",
      y = "Share of county non-correspondence")
```

We can also see that non-correspondence is disproportionately driven by major
events, perhaps because these types of events have initial PDAs submitted early with just a subset of the
most impacted counties, and then the requests are amended over time as the full extent of damages becomes clearer.

Notably, we cannot evaluate whether any of these patterns hold for denied requests since FEMA does not publish any
county-level details in its authoritative request denial dataset.

```{r}
# the top disasters that drive county non-correspondence are disproportionately large-dollar-value events
pdas %>%
  mutate(
    pda_pa_cost_estimate_total_millions = round(pda_pa_cost_estimate_total / 1000000, digits = 0), 
    total_cost_percentile = ntile(pda_pa_cost_estimate_total, n = 100)) %>%
  filter(fema_disaster_number %in% (disaster_coverage %>% slice(1:5) %>% pull(fema_disaster_number))) %>%
  select(pda_pa_cost_estimate_total_millions, total_cost_percentile)
```

Because PDAs uniquely provide county-level information for denied requests, they allow us to look beyond just the
statewide per-capita impact to see how county-level impact distributions relate to which requests are approved/denied. 

```{r}
county_pdas %>%
  filter(fema_decision_year > 2007, fema_decision == "Denied") %>%
  mutate(
    pda_pa_pci_county_ratio = pda_pa_per_capita_impact_county / pda_pa_per_capita_impact_indicator_countywide,
    pda_pa_pci_county_ratio_logged = log(pda_pa_pci_county_ratio)) %>%
  filter(
    ## produced when pda_pa_pci_county_ratio == 0
    !is.infinite(pda_pa_pci_county_ratio_logged)) %>%
  mutate(
    .by = fema_declaration_request_number,
    county_pci_min = min(pda_pa_pci_county_ratio_logged, na.rm = TRUE),
    county_pci_max = max(pda_pa_pci_county_ratio_logged, na.rm = TRUE)) %>%
  filter(
    if_all(
      .cols = c(pda_pa_pci_county_ratio_logged, county_pci_min, county_pci_max),
      .fns = ~ !is.na(.x) & !is.infinite(.x))) %>%
  ggplot(aes(
    x = fema_declaration_request_number, 
    color = pda_pa_pci_county_ratio_logged > 0)) +
    geom_point(aes(y = pda_pa_pci_county_ratio_logged), alpha = .3) +
    geom_errorbar(
      data = . %>% distinct(fema_declaration_request_number, .keep_all = TRUE), 
      aes(x = fema_declaration_request_number, ymin = county_pci_min, ymax = county_pci_max),
      color = "grey",
      alpha = .5) +
    annotate(
      "text", x = 45, y = 2.75, hjust = 1,
      label = str_wrap("Counties above zero exceed the per-capita threshold", 25)) +
    annotate(
      "curve", x = 44, y = 2.1, xend = 45, yend = 0.1, curvature = -0.35,
      arrow = arrow(length = unit(0.02, "npc"), type = "closed")) +
    labs(
      x = "", y = "", 
      title = "Most denials include many counties that exceed the per-capita impact threshold",
      subtitle = str_c(
        "County-level per-capita impacts, logged, for denied declaration requests, 2008-2026.\n",
        "Points represent counties, which are aligned vertically per declaration request.")) +
    coord_cartesian(clip = "off") +
    theme(
      legend.position = "none",
      axis.text.x = element_blank(), 
      axis.ticks.x = element_blank(), 
      axis.line.x = element_blank())
```

## Use of Generative AI

This package was developed in part using agentic AI, under the supervision of the author. 